---
title: 'Assignment 2: Classification'
author: "Joram Keijser"
date: "Wednesday, December 10, 2014"
output: html_document
---

```{r package, include=FALSE}
install.packages("randomForest", repos = "http://cran.xl-mirror.nl/")
library("randomForest")
```

### 0. Introduction

First, download the data. 
```{r download, cache=TRUE, tidy=TRUE}
attributes <- c("class", "handicap", "water", "adoption", "physician", "el-salvador", 
                "religious", "sattelite", "nicaraguan", 
                "missile", "immigration", "synfuels", "education", "superfund", "crime", 
                "dut-free",                 "export")
Data <- read.table("http://archive.ics.uci.edu/ml/machine-learning-databases/voting-records/house-votes-84.data", sep = ',', col.names = attributes, na.strings = "?")
```

There are `r sum(is.na(Data))` missing values. We replace the missing values
by the most occuring value in the particular. For example, if
most votes on the `handicap` topic were a yes, than the missing votes on
this topic will taken to be yes. This is done by the `na.roughfix` function. 
```{r missing, cache=TRUE}
Data <- na.roughfix(Data)
```

```{r meanVotes, include=FALSE, cache=TRUE}
REP <- Data$class == "republican"
DEM <- Data$class == "democrat"
meanVotes <- data.frame(matrix(ncol = dim(Data)[2], nrow = 2))
names(meanVotes) <- attributes
meanVotes[1,1] <- "Republican"
meanVotes[2,1] <- "Democrat"
for( k in 2:dim(Data)[2] ) {
  meanVotes[1, k] <- mean(Data[REP, k] == "y")
  meanVotes[2, k] <- mean(Data[DEM, k] == "y")
}
```
Next, we will make some plots to get a feel for the data. 
The biggest differences in voting behaviour are observed on the topics
Physician, duty-free, education, missile and adoption. 
```{r plots1}
## plot mean number of yes's per party
par(las=2) # make label text perpendicular to axis
par(mar=c(5,8,4,2)) # increase y-axis margin.
barplot(as.matrix(meanVotes[,-1]), beside = TRUE, col = c("red", "blue"), horiz=T, xlim=c(0,1))
#sort(apply(meanVotes, 2, function(x){ max(x)/min(x)}), decreasing = TRUE)
```

Finallly, divide the data in a training and test set.
```{r test/train, cache=TRUE}
trainIndex <- sample.int(dim(Data)[1], size = round(dim(Data)[1]/2))
Train <- Data[trainIndex, ]
Test <- Data[-trainIndex, ]
```

### 1. Naive Bayes classification

```{r NBpackage, include = FALSE}
install.packages("e1071", repos = "http://cran.xl-mirror.nl/")
library("e1071")
```